---
title: "Causal graph"
author: "Iyar Lin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 2
    pandoc_args: --webtex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F, cache = F)
set.seed(1)
options(scipen = 999)

packages <- c(
  "devtools", # install_github
  "tidyverse" # best thing that happend to me
)

sapply(
  packages,
  function(x) if (!require(x, character.only = TRUE, quietly = T)) {
      install.packages(x, quiet = T, verbose = F)
      library(x, character.only = T, quietly = T, verbose = F)
    }
)

if(! require(dagitty)){
  install_github("jtextor/dagitty/r") # Analysis of structural causal models
  library(dagitty, quietly = T)
}

```

# Causal graph

## Variables

$d_t\in \{0,1\}$ - Subscriber disconnected at tenure $t$  

$T \in [0,\infty)$ - Subscriber life time  

comp - Competition measurements  

cong - Network congestion measurements  

dem - Demographics (credit score, population density, age)  

pop - Population count in beam  

pool - Perspective customers pool (count)  

plan - Plan set offered  

subs - Number of subscribers on the network (that share bandwidth)  

t_s - Time since satellite lunched  

cap - Maximum capacity  


```{r plot causal graph}
g <- dagitty( "dag {
d_t -> T
dem -> pool
dem -> comp
pop -> pool
comp -> pool
comp -> plan
comp -> d_t
pool -> subs
plan -> d_t
plan -> subs
plan -> cong
t_s -> subs
subs -> cong
cap -> cong
cong -> d_t
}" )

set.seed(1)
plot(graphLayout(g))
```

A simpler model might be:

```{r plot simpler causal graph}
g <- dagitty( "dag {
d_t -> T
con -> d_t
comp -> con
comp -> d_t
dem -> comp
dem -> d_t
}" )

set.seed(1)
plot(graphLayout(g))
```


## Outages

It's possible that more than congenstion outages causes disconnects. 